---
title: "Viz wages"
author: "Duc-Quang Nguyen | swissinfo.ch"
date: " 2017"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: simplex
---
## TODO
Integrate this [report](http://lenews.ch/2017/05/04/swiss-employees-the-best-paid-in-europe-according-to-study/)

salaire & age
https://www.jobcloud.ch/c/fr-ch/blog/2014/09/barometre-salarial-apres-55-ans-les-augmentations-salaires-se-font-rares-cp/

https://www.forbes.com/sites/learnvest/2014/01/13/your-high-earning-years-salary-secrets-for-your-20s-30s-and-40s/#22ae1f01201e

## TXT

Money can't buy everything. Still, salary might still be an important factor to consider when deciding of a career path. How different occupation differ  We present here at the official European statistics on wages (Earnings Survey 2014)

Switzerland boasts among the highest earning in the world. According to an European survery on salary from 2014, with over 6000 € Swiss' average salary expressed was the highest in Europe as illustrated in the graphic under. This represents more than twice EU-28 mean earnings, 14 times Bulgaria average wage. However expressing the wages in Euros, largely inflates Swiss salaries. Because of the strong Swiss franc, furthermore it doesn't account for the high cost of living. Once standardiszed by purchasing power, Switzerland still tops of European average wage 


https://datawrapper.dwcdn.net/z8s6D/1/

<iframe id="datawrapper-chart-z8s6D" src="//datawrapper.dwcdn.net/z8s6D/1/" frameborder="0" allowtransparency="true" allowfullscreen="allowfullscreen" webkitallowfullscreen="webkitallowfullscreen" mozallowfullscreen="mozallowfullscreen" oallowfullscreen="oallowfullscreen" msallowfullscreen="msallowfullscreen" width="100%" height="600"></iframe><script type="text/javascript">if("undefined"==typeof window.datawrapper)window.datawrapper={};window.datawrapper["z8s6D"]={},window.datawrapper["z8s6D"].embedDeltas={"100":843,"200":675,"300":657,"400":618,"500":600,"600":600,"700":600,"800":600,"900":582,"1000":582},window.datawrapper["z8s6D"].iframe=document.getElementById("datawrapper-chart-z8s6D"),window.datawrapper["z8s6D"].iframe.style.height=window.datawrapper["z8s6D"].embedDeltas[Math.min(1e3,Math.max(100*Math.floor(window.datawrapper["z8s6D"].iframe.offsetWidth/100),100))]+"px",window.addEventListener("message",function(a){if("undefined"!=typeof a.data["datawrapper-height"])for(var b in a.data["datawrapper-height"])if("z8s6D"==b)window.datawrapper["z8s6D"].iframe.style.height=a.data["datawrapper-height"][b]+"px"});</script>


### Wage by occupation 

How do wages differ depending of the occupation? The graphic under presents wage difference for the 8 large groups of occupations for selected european countries. 

These 8 groups of occupations might sound abstract, they are from the International Standard Classification of Occupations. In a nutshell, it is based on the skill level and skill specialisation required for a given job. Regardless of the country considered, most occupations earn wage below country's average wage. Only the three most specialized occupations: Managers, Professional, Technicians and associate professionals are                                                                                             

https://datawrapper.dwcdn.net/jP1tg/1/

<iframe id="datawrapper-chart-jP1tg" src="//datawrapper.dwcdn.net/jP1tg/1/" frameborder="0" allowtransparency="true" allowfullscreen="allowfullscreen" webkitallowfullscreen="webkitallowfullscreen" mozallowfullscreen="mozallowfullscreen" oallowfullscreen="oallowfullscreen" msallowfullscreen="msallowfullscreen" width="100%" height="474"></iframe><script type="text/javascript">if("undefined"==typeof window.datawrapper)window.datawrapper={};window.datawrapper["jP1tg"]={},window.datawrapper["jP1tg"].embedDeltas={"100":600,"200":510,"300":492,"400":474,"500":474,"600":474,"700":474,"800":456,"900":456,"1000":456},window.datawrapper["jP1tg"].iframe=document.getElementById("datawrapper-chart-jP1tg"),window.datawrapper["jP1tg"].iframe.style.height=window.datawrapper["jP1tg"].embedDeltas[Math.min(1e3,Math.max(100*Math.floor(window.datawrapper["jP1tg"].iframe.offsetWidth/100),100))]+"px",window.addEventListener("message",function(a){if("undefined"!=typeof a.data["datawrapper-height"])for(var b in a.data["datawrapper-height"])if("jP1tg"==b)window.datawrapper["jP1tg"].iframe.style.height=a.data["datawrapper-height"][b]+"px"});</script>


 

###


## Snippets


La théorie du capital humain montre que par un investissement continu durant la carrière, les compétences s'accroissent tout au long de la carrière professionnelle et cette expérience accumulée entraîne un salaire croissant dans une perspective de cycle de vie. C'est l'expérience et non l'âge en tant que tel qui est rémunérée.

https://www.jobcloud.ch/c/fr-ch/blog/2014/09/barometre-salarial-apres-55-ans-les-augmentations-salaires-se-font-rares-cp/


Disparities in gross hourly earnings within a country can be measured using deciles, and in particular the lowest and highest deciles, which correspond to the 10% of employees earning the least (D1) and to the 10% earning the most (D9). As a consequence, a high D9/D1 interdecile ratio indicates large disparities. D1 is the maximum gross hourly earnings received by the 10% of employees earnings least. D9 is the minimum gross hourly earnings received by the 10% of employees earning most.



```{r setup, include=FALSE}
iso2.sub <- c('EU28', 'CH', 'DE' , 'FR', 'IT','UK', 'ES', 'AT', 'PT', 'TR')
names(iso2.sub) <- c('#4c6570', '#b15953', rep('#a6a6a6', length(iso2.sub) -2))

eu.monthly.file <- "data/earn_ses_monthly.RData"
load(file = eu.monthly.file)


library(tidyverse)
library(scales)
library(magrittr)
library(countrycode)
library(swiMap)
library(swiTheme)

### Getting data in packages
library(eurostat)

### Interactive 
library(highcharter)
library(htmltools)
library(shiny)
library(ggiraph)
library(swiRcharts)
library(classInt)
library(hrbrthemes)
require(viridis)
```


```{r plot overall average wage,  fig.width= 12, fig.height = 16}

eu.dat$age <- factor(eu.dat$age, levels = c("Y_LT30", "Y30-39", "Y40-49", "Y50-59", "Y_GE60", "TOTAL"))
dd <- eu.dat %>% filter(indic_se == 'ERN_MN_PPS', nace_r2 == 'B-S_X_O', time == 2014) %>% select(-time, -nace_r2, -indic_se)

dd1 <- eu.dat %>% filter(indic_se %in% c('ERN_MN_EUR', 'ERN_MN_PPS', 'ERN_D1_PPS', 'ERN_D9_PPS'), !geo %in% c('EA18', 'EA19'),  age == "TOTAL", nace_r2 == 'B-S_X_O', isco08 == "TOTAL", age == "TOTAL", time == 2014)
dd1 %<>% spread(indic_se, values) %>% mutate(ratio = ERN_D9_PPS / ERN_D1_PPS) %>% 
  select(-nace_r2, -isco08, -age, -time, -ERN_D9_PPS, -ERN_D1_PPS) %>%
  gather(indic_se, values, -geo)

#dd1 %>% mutate()
dd1$group <- ifelse(dd1$geo %in% eurostat::eu_countries[,1], 'EU', 
       ifelse(dd1$geo %in% eurostat::eu_candidate_countries[,1], 'candidate EU', 
              ifelse ( dd1$geo %in% eurostat::efta_countries[,1], 'EFTA', '')
       )
)
dd1 %<>% mutate(geo.label = countrycode(dd1$geo, "eurostat", "country.name")) %>% mutate(geo.label = ifelse(is.na(geo.label), as.character(geo), geo.label))


ggplot(dd1) + geom_bar(aes(x = reorder(geo.label, values), y = values, group=group, fill = group), stat = "identity") + facet_wrap(~ indic_se, scales = "free_x") + 
  coord_flip() + theme_ipsum_rc() + scale_fill_manual(values = swi_col)


dd1.dw <- dd1 %>% select(-geo) %>% spread(indic_se, values) %>% select(2,1, 3:ncol(dd1))
write.csv(dd1.dw, file = "input/averageWage_DW.csv", row.names = F)
```

```{r average wage by occupation, fig.width= 12, fig.height = 16}
#dd2 <- dd %>% group_by(geo, age) %>% mutate(tot.isco = values[which(isco08 == "TOTAL")]) %>% ungroup() %>% filter(isco08 != "TOTAL")
dd2 <- eu.dat %>% filter(indic_se %in% c('ERN_MN_PPS', 'ERN_MN_EUR'), nace_r2 == 'B-S_X_O', time == 2014) %>% 
  select(-time, -nace_r2)  %>% spread(indic_se, values) 

dd2 %<>% group_by(geo, age) %>% mutate(
  `tot.isco.eur` = ERN_MN_EUR[which(isco08 == "TOTAL")], 
  val.eur = round((( ERN_MN_EUR - `tot.isco.eur`) / `tot.isco.eur`) * 100, 1), 
  `tot.isco.pps` = ERN_MN_PPS[which(isco08 == "TOTAL")],
  val.pps = (( ERN_MN_PPS - `tot.isco.pps`) / `tot.isco.pps`) * 100
) %>% ungroup()
all.equal(dd2$val.eur, dd2$val.pps)
dd2 %<>% select(-ERN_MN_PPS, -tot.isco.pps, -val.pps, -tot.isco.eur)


dd2 %<>% filter(geo %in% iso2.sub) %>% mutate(geo = factor(geo, levels = iso2.sub))
dd2$isco08.full <- factor(isco.dic[match(dd2$isco08, isco.dic$code_name), 2] %>% unlist(use.names = F), 
                         levels = isco.dic[match(levels(dd2$isco08)[levels(dd2$isco08) %in% dd2$isco08], isco.dic$code_name), 2] %>% 
                           unlist(use.names = F))

ggplot(dd2 %>% filter(age == "TOTAL")) + geom_tile(aes(x = geo, y = isco08.full, fill = val.eur)) + theme_ipsum_rc() + 
  scale_fill_gradient2(high = "#4c6570", mid = , low = "#72302f", midpoint = 0)



dd2a <- dd2 %>% filter(age == "TOTAL", isco08.full != "Total") %>% mutate(isco08.full = droplevels(isco08.full)) %>% mutate(geo = as.character(geo))
dd2a$geo[which(dd2a$geo == 'UK')] <- 'GB'

dd2a %<>% mutate(geo.label = countryTranslation(ifelse(nchar(dd2a$geo) == 2, dd2a$geo, NA), "EN")[,-1])
dd2a$geo.label <- ifelse(is.na(dd2a$geo.label), dd2a$geo, dd2a$geo.label)
dd2a$geo.label <- factor(dd2a$geo.label, levels = dd2a[match(gsub("UK", "GB", iso2.sub), dd2a$geo), "geo.label"] %>% unlist(use.names = F))


#dd2a %<>% mutate(geo.label = countrycode(geo, "iso2c", "country.name")) %>% mutate(geo.label = ifelse(is.na(geo.label), geo, geo.label))

dd2a$tooltip <- paste0(
    '<table cellpadding="1" style="line-height:0.9">',
    '<tr><td><strong>', dd2a$geo.label,'</strong></td></tr>',
    '<tr><td><i>', dd2a$isco08.full, '</i></td></tr>',
    '<tr><td>', "Average wage", ': <strong>', dd2a$ERN_MN_EUR, ' €</strong>', '</td></tr>',
    '<tr><td>', "% difference with country's average wage", ": <strong>", dd2a$val.eur, '%</strong></td></tr>',
    '</table>')

lang <- 'EN'
# make an interactive version 
hc <- hchart(dd2a, "heatmap", 
             hcaes(x = geo.label, y = isco08.full, value = val.eur, tooltip = tooltip)) %>% 
  hc_chart(marginRight = 2) %>% 
  hc_colorAxis( stops = list(list(0,  swi_col[length(swi_col)]), list(0.33, "white"), list(1, swi_col[1])),  min = -50, max = 100) %>%
  hc_legend(layout = "horizontal", verticalAlign = "top", width =  90, padding = 0,
            align = "center", valueDecimals = 0, x = -10, rtl = ifelse(lang == "AR", T, F), 
            reversed = ifelse(lang == "AR", T, F)) %>%
  hc_add_theme(hc_theme_swi) %>%
  hc_yAxis(reversed = TRUE, offset = 2, tickLength = 0, lineWidth = 0,
           gridLineWidth = 0, minorGridLineWidth = 0,
           title = list(text = ""), labels = list(y = 4, style = list(fontSize = "12px")),
           opposite = ifelse(lang == "AR", T, F)) %>%
  hc_xAxis(lineWidth = 0, title = list(text = ""), reversed = ifelse(lang == "AR", T, F), opposite = TRUE) %>% 
  hc_size(height = 530) %>% 
  hc_tooltip(
    formatter = JS("function() { return this.point.tooltip;}"), 
    useHTML = TRUE,
    borderWidth = 1,
    padding = 3
  )

hc


#footer <- paste0(txt['source', lang], ": ", txt['source.name', lang])

save_html(
  tags$html(
    tags$head(includeHTML(style_swi_highcharter())),
    tags$body(    
     # tags$h2(txt['main.title', lang]),
     # div(class = "descr", HTML(txt['descr', lang])),
     # tags$h3(txt['h3', lang]),
      div(class="graphic", hc),
      #div(id = "cite", HTML(footer)),
      HTML(iframeresizer)
    )
  ), file = paste0("heatmap_wageByISCO08_", lang, ".html"), libdir = "js"
)     

  
```


```{r salary by age}

dd3 <- eu.dat %>% 
  filter(indic_se == 'ERN_MN_EUR', nace_r2 == 'B-S_X_O', time == 2014, geo %in% iso2.sub) %>% 
  select(-time, -nace_r2, -indic_se)

dd4 <- dd3 %>% arrange(age) %>% group_by(isco08, geo) %>% 
  mutate(diff = c(NA, (values[-1] - values[-c(length(values))] ) / values[-c(length(values))] )) %>%
  ungroup() %>% 
  filter(!age %in% c('TOTAL', 'Y_LT30'))

dd4.p <- ggplot(dd4 %>% filter(isco08 != "TOTAL")) + geom_bar(aes(x = age, y = diff, fill = age), stat = "identity") +
  facet_grid(geo ~ isco08) + theme_ipsum_rc() +
  scale_fill_ipsum() 


dd3 %<>% group_by(isco08, geo) %>% arrange(age) %>% 
  mutate(wage.lt30 =  values[1]) %>% ungroup() %>% 
  filter(age != "TOTAL")

dd3 %<>% mutate(val = (values - wage.lt30) / wage.lt30) 
                
dd3$isco08.full <- factor(isco.dic[match(dd3$isco08, isco.dic$code_name), 2] %>% 
  unlist(use.names = F), levels = isco.dic[match(levels(dd3$isco08), isco.dic$code_name), 2] %>% unlist(use.names = F))

## add colour, line size
dd3 %<>% mutate(color = names(iso2.sub)[match(geo, iso2.sub)], size = ifelse(geo %in% c('CH', 'EU28'), 0.5, 0.1))


# dd3.gp <- ggplot(dd3 %>% filter(isco08 != "TOTAL")) + 
#   geom_line(aes(x = age, y = val, group = geo, colour = geo, size = size), alpha = 0.7) +
#   geom_point(aes(x = age, y = val, group = geo, colour = geo, size = size)) + 
#   facet_wrap(~isco08.full, nrow = 3) + swi_theme() +
#   theme() + 
#   scale_fill_gradient2(high = "#152c37", mid = "white", low = "#72302f", midpoint = 0) + geom_hline(yintercept = 0) +
#   scale_colour_ipsum() + scale_size_identity() + ggtitle("% difference in average wage by age and occupation groups") + ylim(c(0, 100))
# 
# dd3.gp
# 
# dd3i <- ggplot(dd3 %>% filter(isco08 != "TOTAL")) + 
#   geom_line(aes(x = age, y = val, group = geo, colour = geo, size = size, data_id = interaction(age, geo, isco08.full)), alpha = 0.7) +
#   geom_point_interactive(aes(x = age, y = val, group = geo, colour = geo, size = size)) + 
#   facet_wrap(~isco08.full, ncol = 2) + swi_theme() +
#     theme(
#       legend.position = "none", 
#      # plot.background = element_rect(fill = "#333333", size = 0, color = NA),
#       panel.grid.major.y=element_line(color="#737373", linetype="dotted", size=0.09),
#      # axis.title = element_text(colour = "#f7f5ed", size = 15),
#       #axis.text = element_text(colour = "#e5dbcd"),
#       strip.background = element_blank(),
#       strip.text = element_blank(),
#       plot.margin=unit(c(0, 0.1, 0, 0.1), "cm")) +
#   scale_fill_gradient2(high = "#152c37", mid = "white", low = "#72302f", midpoint = 0) + geom_hline(yintercept = 0) +
#   scale_size_identity() + ylim(c(0, 100))
# 
# 
# ggiraph(code = {print(dd3i)})

lang <- 'FR'
oc <- "OC1"
charts <- lapply(unique(dd3$isco08)[unique(dd3$isco08) != "TOTAL"], function(oc) {
 
  ageGroup <-  c(">30 ans", "30-39", "40-49 ", "50-59", "<60 ans")
    
  # get different translations
  dddd <- dd3 %>% filter(isco08 == oc) %>% select(-wage.lt30) %>% 
    filter(age != 'Y_LT30') %>%
    mutate(geo = ifelse(as.character(geo) == 'EU28', 'EU', as.character(geo)),
           geo = ifelse(as.character(geo) == 'UK', 'GB', geo)) %>%
    mutate(label = countryTranslation(geo, lang)[,2]) %>%
    # get age translation
    mutate(age = factor(age, labels = ageGroup[-1]))

  dddd$tip <- paste0(
    '<h4>', as.character(dddd$label), '</h4><div class="tpsubtitle">',
    as.character(dddd$isco08.full), '</div><div class = "tp">',
    "Salaire moyen des ", as.character(dddd$age), " comparé aux ", ageGroup[1], 
    " :<b>", ifelse(dddd$val >0, "+", "-"), round(dddd$val, 2) * 100, "</b>%<br>",
    "Salaire moyen: ", dddd$values, '€</div>')
  dddd$tip <- gsub("'", "_", gsub("\\\n", "", dddd$tip))
  
  gp <- ggplot(dddd, aes(x = age, y = val, group = geo, colour = color)) + 
  geom_line(aes(size = size + 0.1), alpha = 0.4) +
  geom_point_interactive(
    aes(size = size + 1.5, alpha = size + 0.1, 
        data_id = interaction(age, geo, isco08.full), tooltip = tip)) + 
  scale_size_identity() +
  scale_colour_identity() +  
  scale_y_continuous(limits = c(0, 1), labels=percent, expand = c(0.01, 0), name = "") +
  scale_x_discrete( name = "", expand = c(0.02, 0.02)) 
  
  if(oc %in% c("OC1", "OC2")) {
    gp <- gp + geom_text(data = dddd %>% filter(geo %in% c('CH', 'EU'), age == ageGroup[length(ageGroup)]), 
                   aes(label = label), hjust = 0.9)
  }
  
  gpl <- gp +  swi_theme(base_size = 12, base_family = "Open Sans Condensed", base_color = "#152c37") +
    theme(
      legend.position = "none", 
     # plot.background = element_rect(fill = "#333333", size = 0, color = NA),
      panel.grid.major.y=element_line(color="#8c8c8c", size=0.02),
      axis.ticks.x = element_line(size = 1, colour = '#595959'),
      axis.text.y = element_text(margin=margin(r=-4)),
      axis.line.x = element_line(size=0.5),
     # axis.title = element_text(colour = "#f7f5ed", size = 15),
      #axis.text = element_text(colour = "#e5dbcd"),
      #plot.title=element_text(margin=margin(b=3)),
      #strip.background = element_blank(),
      #strip.text = element_blank(),
      plot.margin=unit(c(0.5, 0.6, .3, 0), "cm"))
    
  gpli <- ggiraph(code = {print(gpl)},
    hover_css = "fill-opacity:0.6;stroke-opacity:0.9;stroke-width:1.1px;stroke:#333333;",
    tooltip_opacity = 0.7,
    pointsize = 12,
    width = 1,
    width_svg = 3.7,
    height_svg = 3.3,
    fonts = list(sans = "Open Sans Condensed", serif = "Open Sans Condensed"))
    list(chart = gpli, title = as.character(unique(dddd$isco08.full)))
})





save_html(
   tags$html(
      tags$head(includeHTML("styles.html")),
  fluidPage(
    #tags$h2(txt["main.title", lang]),
    #div(class = "descr", HTML(paste0(txt["descr", lang], "<br>", txt["descr2", lang]))),
    #tags$h4(txt["h3", lang]),
    div(class="graphic", 
        fluidRow(lapply(1:length(charts), function(i) {
          column(4, div(class = "graphic",  h4(HTML(charts[[i]]$title)), charts[[i]]$chart))
        }))#,
        #div(id = "cite", HTML(footer))
    ),
    HTML(iframeresizer)  
  )), #background = "#EDEDED", 
  file =  paste0("wageByISCOandAge_", lang, ".html"), libdir = "js")
  




```